Appendix: Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-06 appendix.qmd
Make the plan for the methodology

$$
s = \frac{r_2}{r_1}
$$ {#eq-equation2}
-->

>>>>

$$
s = \frac{r_2}{r_1}
$$ {#eq-equation2}

<!--UMAP parm 2-->

## Import data 
training_data_pbmc <- read_rds("data/pbmc/pbmc_3k_festem/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:15] |>
  mutate(ID = 1:NROW(training_data_pbmc))

UMAP_pbmc <- read_rds("data/pbmc/pbmc_umap_15_min_dist_0.99_metric_cosine.rds")

class_avg <- UMAP_pbmc |>
  group_by(cell_label) |>
  summarise(
    UMAP1 = median(UMAP1),
    UMAP2 = median(UMAP2)
  ) |>
  mutate(cell_label = as.numeric(cell_label)) |>
  arrange(cell_label)

levels(class_avg$cell_label) <- c("Memory \nCD4 T", "Naive CD4 T", "CD14+ Mono",  "B", "CD8 T", "FCGR3A+ \n Mono", "NK", "M-MDSC\n-like", "CD27-CD+ \n Memory T", "DC")

plot_list2_pbmc <- UMAP_pbmc |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = cell_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) 
```

shape_value_pbmc <- calculate_effective_shape_value(.data = UMAP_pbmc,
                                                      x = UMAP1, y = UMAP2)

num_bins_vec <- 1:35 ## Number of bins along the x-axis

  model_object <- fit_high_d_model(training_data = training_data_pbmc, nldr_df_with_id = UMAP_pbmc, x = "UMAP1", y = "UMAP2", num_bins_x = num_bins_vec[i], shape_val = shape_value_pbmc,
                             is_bin_centroid = TRUE,
                             is_rm_lwd_hex = FALSE,
                             benchmark_to_rm_lwd_hex = NA,
                             is_avg_high_d = TRUE, column_start_text = "PC_")

  pred_df_training <- predict_2d_embeddings(test_data = training_data_pbmc, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "UMAP")

  eval_df_training <- generate_eval_df(data = training_data_pbmc, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "PC_")

MSE_df_pbmc <- eval_data_training |>
  dplyr::mutate(data_type = "training",
                method = "UMAP")
```

## To draw with AIC
aic_plot_pbmc <- ggplot(MSE_df_pbmc |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 192, linetype="dashed", 
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot_pbmc <- ggplot(MSE_df_pbmc |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 192, linetype="dashed", 
                color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("total number of bins")

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 144, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltPBMC2
##| out-width: 100%
#| fig-pos: H

aic_plot_pbmc + mse_plot_pbmc +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

```{r}
#| warning: false
#| echo: false
shape_value_pbmc <- calculate_effective_shape_value(.data = UMAP_pbmc,
                                                      x = UMAP1, y = UMAP2) ## 0.8772751

# num_bins_pbmc <- find_non_empty_bins(nldr_df = UMAP_pbmc, x = "UMAP1", y = "UMAP2", shape_val = shape_value_pbmc, non_empty_bins = num_non_empty_bins_pbmc)
num_bins_pbmc <-  11

## To extract bin centroids
hexbin_data_object_pbmc <- extract_hexbin_centroids(nldr_df = UMAP_pbmc,
                                                      num_bins = num_bins_pbmc,
                                                      shape_val = shape_value_pbmc, x = UMAP1, y = UMAP2)

df_bin_centroids_pbmc <- hexbin_data_object_pbmc$hexdf_data

UMAP_pbmc_with_hb_id <- UMAP_pbmc |>
  dplyr::mutate(hb_id = hexbin_data_object_pbmc$hb_data@cID)

## To generate a data set with high-D and 2D training data
df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), UMAP_pbmc_with_hb_id)

## Averaged on high-D
df_bin_pbmc <- avg_highD_data(.data = df_all_pbmc, column_start_text = "PC")

## Triangulate bin centroids
tr1_object_pbmc <- triangulate_bin_centroids(df_bin_centroids_pbmc, x, y)
tr_from_to_df_pbmc <- generate_edge_info(triangular_object = tr1_object_pbmc)

## Compute 2D distances
distance_pbmc <- cal_2d_dist(.data = tr_from_to_df_pbmc)

# distance_plot <- plot_dist(distance_pbmc) +
#   #ggtitle("(b)" ) +
#   ylab(expression(d^{(2)})) +
#   theme(axis.text = element_text(size = 5),
#         axis.title = element_text(size = 12))
# 
# distance_plot

## To find the benchmark value
benchmark_pbmc <- find_benchmark_value(.data = distance_pbmc, distance_col = distance)
#benchmark_pbmc <- 1.049766 
benchmark_pbmc <- 2.5

trimesh_removed_pbmc_umap <- remove_long_edges(.data = distance_pbmc, benchmark_value = benchmark_pbmc,
                                                 triangular_object = tr1_object_pbmc, distance_col = distance)

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  geom_point(size = 0.005, colour = "#33a02c") +
  #ggtitle("(b)") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

df_bin_train <- df_bin_pbmc
names(df_bin_train)[-1] <- paste0("avg_", names(df_bin_train)[-1])

error_df <- df_all_pbmc |>
  dplyr::left_join(df_bin_train, by = c("hb_id" = "hb_id")) ## Map high-D averaged/weighted mean coordinates

  error_df[ , paste0("abs_residual_", "PC_", i)] <- abs(error_df[ , paste0("PC_", i)] - error_df[ , paste0("avg_", "PC_", i)])

error_df <- error_df |>
  dplyr::mutate(total = rowSums(dplyr::pick(tidyselect::starts_with(paste0("abs_residual_", "PC_")))))

# library(ggbeeswarm)
# error_df$group <- "1"
# ggplot(error_df, aes(x = group, y = total)) +
#   geom_quasirandom()+
#   ylim(0, max(unlist(error_df$total))+ 0.5) + coord_flip()

### The minimum error is 0 and the maximum is 42.17439
### There is lot of points with error 0,

error_df <- error_df |>
  mutate(type = if_else(total <= 2, "error 2 or less",
                        if_else(total <= 10, "error 2-10",
                                if_else(total <= 15, "error 10-15",
                                        if_else(total <= 20, "error 15-20",
                                                if_else(total <= 25, "error 20-25",
                                                        if_else(total <= 30, "error 25-30",
                                                                if_else(total <= 35, "error 30-35", "error greter than 35"))))))))

plot_list2_pbmc_error <- error_df |>
    mutate(type = factor(type , levels = c("error 2 or less", "error 2-10", "error 10-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error greter than 35"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d"))

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 15, min_dist = 0.99, metric = cosine) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<>)"
#| label: fig-nldervisPBMCUMAP2
#| out-width: 100%
#| fig-pos: h

plot_list2_pbmc + trimesh_removed_pbmc_umap + plot_list2_pbmc_error +
plot_layout(guides='collect', ncol=3) &
  theme(legend.position='bottom')
```
-->
